A Decision Procedure for String Constraints with String-Integer Conversion and Flat Regular Constraints

### 2021-05-15 FMCAD2021/3_flatten.tex
section 4
some change on sec2
sec6 half done
optimization

In $\strint$, there are two types of variables, namely, the string variables $\svarx,\svary,\ldots \in \svars$ and the integer variables $\ivarx,\ivary,\ldots \in \ivars$.
%
A $\strint$ formula $\varphi$ is defined by the following rules, where $\len(\sterm)$ denotes the length of a string $\sterm$.
\[
\begin{array}{l c l}
\sterm & ::= & a \mid \svarx \mid \sterm \concat \sterm, \\
\iterm & ::= & n \mid \ivarx \mid \len(\sterm) \mid \parseInt(\sterm) \mid \iterm + \iterm \mid \iterm - \iterm,\\
\varphi & ::= & \sterm = \sterm \mid x \in \Aut \mid \iterm\ \op\ \iterm \mid \varphi \wedge \varphi \mid \varphi \vee \varphi \mid \neg \varphi,
\end{array}
\]
where $a \in (\Sigma_{\textit{num}})_\varepsilon$, $n \in \Nat$, $\Aut$ is an FA, and $\op \in \{=, < , >, \le, \ge\}$. Let us call $\sterm$ as string terms, $\iterm$ as integer terms, $\sterm = \sterm$ as string equality constraints, $x \in \Aut$ as regular constraints, $\iterm \op\ \iterm$ as arithmetic constraints. 
Let  $\svar(\varphi)$ and $\ivar(\varphi)$ denote the set of string variables and integer variables respectively occurring in $\varphi$.

@@

\subsection{An Extension of Presburger Arithmetic with Exponential Functions ($\paexp$)}

@@

The syntax of {\paexp} is obtained from that of {\pa} by adding $10^\iterm$, $\ell_{10}(\iterm)$ and $\lambda_{10}(\iterm)$ to the definition of integer terms. Specifically, {\paexp} formulas are defined by the rules,
%
 $$
 \begin{array}{ l c l}
 \iterm &::=& c \mid \ivarx \mid \iterm + \iterm \mid \iterm - \iterm \mid 10^\iterm \mid \ell_{10}(\iterm) \mid \lambda_{10}(\iterm), \\
 \phi &::= & \iterm \ \op\ 0 \mid c | \iterm \mid \phi \wedge \phi \mid \phi \vee \phi \mid \neg \phi \mid \exists \ivarx.\ \phi \mid \forall \ivarx.\ \phi.
 \end{array}
 $$
 The semantics of {\paexp} are defined similarly as {\pa}.

@@

\begin{example}
Suppose $\parseInt(\svarx) = 2\ivarx$ is an atomic constraint and $\flatfun_\varphi(\svarx) = (12)^*(34)^*$. Then 
\[
\small
\begin{array}{l l}
& \flatten_{\flatfun_\varphi}(\parseInt(\svarx)  =  2\ivarx)  \\
\equiv & \parseInt(12) \frac{10^{2\#_{\svarx,1}}-1}{10^2-1}10^{2\#_{\svarx,2}}  + \parseInt(34) \frac{10^{2\#_{\svarx,2}}-1}{10^2-1} = 2\ivarx   \\
\equiv & 12 (10^{2\#_{\svarx,1}+2\#_{\svarx,2}} - 10^{2\#_{\svarx,2}})  + 34 (10^{2\#_{\svarx,2}}-1) = 198\ivarx.
\end{array}
\]
%Take $n={\parseInt}((11)^a(10)^b)$ for example.
%\begin{align}
%    n=& {\parseInt}((11)^a)\cdot 2^{2b} + {\parseInt}((10)^b) \notag \\
%    =& {\parseInt}(11)\cdot \frac{2^{2a}-1}{2^2-1}\cdot 2^{2b} + 
%    {\parseInt}(10) \cdot \frac{2^{2b}-1}{2^2-1} \notag 
%\end{align}
\end{example}

>>>>

In $\strint$, there are two types of variables, namely, the string variables $\svarx,\svary,\ldots \in \svars$ and the integer variables $\ivarx,\ivary,\ldots \in \ivars$.
%
A $\strint$ formula $\varphi$ is defined by the following rules, where $\len(\sterm)$ denotes the length of a string $\sterm$.
\[
\begin{array}{l c l}
\sterm & \Def & a \mid \svarx \mid \sterm \concat \sterm, \\
\iterm & \Def & n \mid \ivarx \mid \len(\sterm) \mid \parseInt(\sterm) \mid \iterm + \iterm \mid \iterm - \iterm,\\
\varphi & \Def & \sterm = \sterm \mid x \in \Aut \mid \iterm\ \op\ \iterm \mid \varphi \wedge \varphi \mid \varphi \vee \varphi \mid \neg \varphi,
\end{array}
\]
where $a \in (\Sigma_{\textit{num}})_\varepsilon$, $n \in \Nat$, $\Aut$ is an FA, and $\op \in \{=, < , >, \le, \ge\}$. Let us call $\sterm$ as string terms, $\iterm$ as integer terms, $\sterm = \sterm$ as string equality constraints, $x \in \Aut$ as regular constraints, $\iterm \op\ \iterm$ as arithmetic constraints. 
Let  $\svar(\varphi)$ and $\ivar(\varphi)$ denote the set of string variables and integer variables respectively occurring in $\varphi$.

@@

\subsection{An Extension of Presburger Arithmetic with Exponential Functions (\paexp)}

@@

The syntax of {\paexp} is obtained from that of {\pa} by adding $10^\iterm$, $\ell_{10}(\iterm)$ and $\lambda_{10}(\iterm)$ to the definition of integer terms. Specifically, {\paexp} formulas are defined by the rules,
%
 $$
 \begin{array}{ l c l}
 \iterm &\Def& c \mid \ivarx \mid \iterm + \iterm \mid \iterm - \iterm \mid 10^\iterm \mid \ell_{10}(\iterm) \mid \lambda_{10}(\iterm), \\
 \phi &\Def & \iterm \ \op\ \iterm \mid c | \iterm \mid \phi \wedge \phi \mid \phi \vee \phi \mid \neg \phi \mid \exists \ivarx.\ \phi \mid \forall \ivarx.\ \phi.
 \end{array}
 $$
 The semantics of {\paexp} are defined similarly to {\pa}. The quantifier-free as well as existential {\paexp} formulas are defined similarly to {\pa} as well.

@@

\begin{example}
Suppose $\parseInt(\svarx) = 2\ivarx$ is an atomic constraint and $\flatfun_\varphi(\svarx) = 1^*2^*$. Then 
\[
\small
\begin{array}{l l}
& \flatten_{\flatfun_\varphi}(\parseInt(\svarx)  =  2\ivarx)  \\
\Def & 1 \frac{10^{\#_{\svarx,1}}-1}{10-1}10^{\#_{\svarx,2}}  + 2 \frac{10^{\#_{\svarx,2}}-1}{10-1} = 2\ivarx   \\
\equiv & 10^{\#_{\svarx,1}+\#_{\svarx,2}} - 10^{\#_{\svarx,2}}  + 2 (10^{2\#_{\svarx,2}}-1) = 18\ivarx \\
\equiv & 10^{\#_{\svarx,1}+\#_{\svarx,2}} +  10^{\#_{\svarx,2}} = 18\ivarx + 2.
\end{array}
\]
%Take $n={\parseInt}((11)^a(10)^b)$ for example.
%\begin{align}
%    n=& {\parseInt}((11)^a)\cdot 2^{2b} + {\parseInt}((10)^b) \notag \\
%    =& {\parseInt}(11)\cdot \frac{2^{2a}-1}{2^2-1}\cdot 2^{2b} + 
%    {\parseInt}(10) \cdot \frac{2^{2b}-1}{2^2-1} \notag 
%\end{align}
\end{example}

### 2021-05-15 FMCAD2021/5_optimization.tex
optimization

@@

\paragraph{modified \textbf{Elim-exp}}

>>>>

In the last section, we illustrated the main idea of the decision procedure for  {\paexp}. The decision procedure has a high complexity in the sense that the elimination of the exponential occurrences   of each variable incurs an exponential blow-up, similarly for linear occurrences. Note that this high complexity holds even for quantifier-free formulas containing exponential terms: For a quantifier-free formula $\varphi$ containing exponential terms, we solve its satisfiability problem by adding the existential quantifiers for all the variables occurring in $\varphi$, then eliminate the quantifiers one by one, resulting into $\ltrue$ or $\lfalse$ in the end. The original formula $\varphi$ is satisfiable if $\ltrue$ is obtained in the end.

In this section, we focus on quantifier-free {\paexp} formulas (or existential {$\paexp$} formulas since they are satisfiability-equivalent), and present various optimizations of the decision procedure for {\paexp}, aiming at an efficient implementation. The focus on quantifier-free {\paexp} formulas is motivated by the following two facts: 1) the flattening of {$\strint$} string constraints results into such formulas, 2) these formulas are already challenging for the state-of-the-art SMT solvers e.g. CVC4 and Z3 (with exponential functions defined as recursive functions).

Let $\varphi$ be a quantifier-free {\paexp} formula in the remainder of this section. Moreover, we assume that $\varphi$ is normalized since the optimizations presented in the sequel are for normalized formulas.

\paragraph{Reducing the number of enumerated variable orders by over approximation}

Recall that in the decision procedure for {\paexp}, after the normalization, the orders of variables are enumerated and for each order, the exponential as well as linear occurrences of variables are eliminated. Since the quantifier elimination is expensive and applied to each possible order of variables, if we could reduce the candidate variable orders in the very beginning, it would facilitate considerable speed-up for the decision procedure.

Our main idea is to consider an over approximation of $\varphi$, which is a {\pa} formula $\varphi'$, and use $\varphi'$ to remove the infeasible candidate variable orders.

Note that all the exponential terms in $\varphi$ is of the form $10^{\ivarz}$ for some integer variable $\ivarz$. 
%
The over approximation is based on the observation that $10^n \ge 9 n + 1$ for every $n \in \Nat$. Then we obtain the over approximation $\varphi'$ from $\varphi$ by replacing each exponential term $10^{\ivarz}$ with a fresh variable $\ivarz'$ and add $\ivarz' \ge 9 \ivarz +1$ as a conjunct.

Then during the enumeration of the linear orders for the variables $\ivarx_1, \ldots, \ivarx_n$, we can quickly remove those infeasible candidates $\sigma$ such that $\varphi' \wedge \bigwedge \limits_{i \in [n-1]}\ivarx_{\sigma(i)} \le \ivarx_{\sigma(i+1)}$ is unsatisfiable. A special case is that if $\varphi'$ is unsatisfiable, then we can directly conclude that original formula $\varphi$ is unsatisfiable.

%Given any formula in $\paexp$, we first invoke \textbf{Normalization} to translate it into a simpler form. Then we  try to detect if there are obvious conflicts, for example, if the linear constraints combined are unsatisfiable, the whole formula is of course unsatisfiable.

%Suppose the normalized formula $\theta$ contains variables $\{x_i:i\in[n]\}$, so the exponential terms could be $e^{x_i}$. For each $e^{x_i}$, we subtitute $e^{x_i}$ by a fresh variable $y_i$ in $\theta$ and add the constraints $(e-1)x_i+1\le y_i$ to get $\theta'$. Since $(e-1)x_i+1\le e^{x_i}$ holds for all $e,x\in \Nat$, $\theta'$ is an over-approximation of $\theta$ and if $\theta'$ is unsatisfiable over positive reals, then $\theta$ is unsatisfiable over $\Nat$.

\zhilin{stopped here}

\paragraph{Avoid the elimination of linear occurrences of variables by under approximation}

xxx

\paragraph{Avoid the formula-size blow-up by depth-first search}

xxx

\paragraph{pruning and preprocessing}

xxx

@@

\paragraph{modified \textbf{Elim-exp}}\label{para: modified Elim-exp}
